---
title: "ECN 102: Analysis of Economics Data"
subtitle: "Chapter 10: Multivariate Regression"
author: "Remy Beauregard"
lang: en
format:
  revealjs:
    output-file: "ECN 102 Lecture 10.html"
    slide-level: 2
    theme: simple
    transition: none
    smaller: true
    include-in-header:
      text: |
        <meta name="description" content="Lecture slides 10 for ECN 102.">
  beamer:
    output-file: "ECN 102 Lecture 10.pdf"
  html:
    output-file: "ECN 102 Lecture 10 Plain.html"
    include-in-header:
      text: |
        <meta name="description" content="Lecture slides 10 for ECN 102.">
execute:
  echo: false
---

```{r}
#| label: setup
#| include: false
require("Statamarkdown")
```

## Motivation for Multivariate Regression

Previously, we modeled and conducted inference on the relationship between two variables, X and Y. However, we are likely to encounter many scenarios where we think more than one explanatory variable is important for explaining an outcome variable. In this case, we need to utilize multivariate regression.

## Challenges of Multivariate Regression

Multivariate regression is a bit more complicated than bivariate:

1) We cannot produce a scatterplot in more than 2 (or 3) dimensions, meaning we may not easily be able to visualize our series together in a single plot.

2) We also cannot compute correlation between more than two variables, meaning our statistics will be a bit more complicated.

3) We may think that some regressors covary with others, meaning we will need to work harder to isolate the unique contribution each regressor makes to changes in our outcome variable.

## Pairwise Correlations

We may still want to think about correlation between variables in the multivariate case, but we are restricting to two variables at a time. Instead, we may compute and display a set of **pairwise correlations**: how each variable of interest varies with all other variables individually. We have seen an example of this in Stata:

```{stata}
#| label: load-auto
#| collectcode: true
qui sysuse auto
```

```{stata}
#| label: corr-auto
corr price mpg weight rep78
```

## The Multivariate Regression Model

Our multivariate regression model will look quite similar to the bivariate model but with additional terms:
$$\hat{y}=b_1+b_2x_2+b_3x_3+...+b_kx_k$$

where $\hat{y}$ is our outcome or *dependent* variable and $x_2...x_k$ are our $k-1$ independent variables or *regressors*. $b_2...b_k$ are then the estimated coefficients associated with each regressor.

Our $i^{th}$ observation will be given as $\hat{y}_i=b_1+b_2x_{2i}+...+b_kx_{ki}$.

## Residuals in Multivariate Regression

Just like bivariate regression, our *residual* is the vertical distance between our actual and fitted values: $e_i=y_i-\hat{y}_i$, which can now be expanded to
$$e_i=y_i-\underbrace{b_1-b_2x_{2i}-b_3x_{3i}-...-b_kx_{ki}}_{\hat{y}_i}$$

However, we need to know how to estimate the $k$ unknown coefficients in our regression model, $b_1,...,b_k$ 

## Normal Equations (First-Order Conditions)

Our $k$ OLS coefficients are found as the solution to $k$ first order conditions or **normal equations** set equal to zero:
$$\sum_{i=1}^nx_{1i}(y_i-b_1-b_2x_{2i}-...-b_kx_{ki})=0$$
$$\sum_{i=1}^nx_{2i}(y_i-b_1-b_2x_{2i}-...-b_kx_{ki})=0$$
$$...$$
$$\sum_{i=1}^nx_{ki}(y_i-b_1-b_2x_{2i}-...-b_kx_{ki})=0$$

We obtain these by applying the chain rule to $\frac{\partial}{\partial b_j}\sum_i e_i^2$ for $j=1,...,k$.

## Orthogonality

Given our definition of our residual as $e_i=y_i-\hat{y}_i$ and our normal equations above, we can rewrite these FOCs as
$$\sum_{i=1}^nx_{ji}e_i=0\text{ for }j=1,...,k$$
We see that this means the sum of the products of each regressor and the residual is zero, also called **orthgonality**.

Since we know that our constant term means $x_{1i}=1$ for all $i$, we can show that $\sum_ie_i=0$ for OLS:
$$\sum_{i=1}^nx_{1i}e_i=0\rightarrow\sum_{i=1}^n1e_i=0\rightarrow\sum_{i=1}^ne_i=0$$

## Perfect Collinearity

An additional constraint on our multivariate regression is that our regressors are not **perfectly colinear** with each other (or with linear combinations of each other). For example, if we had a regressor $x_2=3x_3$, we could not include both $x_2$ and $x_3$ in our multivariate regression.

Even if our regressors are not *perfectly* colinear, we may still be concerned if two or more of our regressors are highly (but not perfectly) correlated.

Since we are solving $k$ FOCs to find $k$ unknown coefficients, we also need to ensure that our sample size exceeds our number of regressors: $n>k$

## Estimating Coefficients: The Formula for $b_j$

Unlike bivariate regression, we cannot easily express a complete formula for our coefficient estimates.

If we consider $b_j$ to be the coefficient on the $j^{th}$ regressor in a multivariate regression, our coefficient will be equal to the ratio of the covariance (with our residual) to the variance of *the residual of a regression of $x_j$ on all other regressors*:

$$b_j=\frac{\sum_{i=1}^n\tilde{x}_{ji}(y_i-\bar{y})}{\sum_{i=1}^n\tilde{x}_{ji}^2}=\frac{Cov(\tilde{x}_j,y)}{Var(\tilde{x}_j)}$$

where our residual $\tilde{x}_{ji}=x_{ji}-\hat{x}_{ji}$ and $\hat{x}_{ji}$ is the predicted value of $x_{ji}$ from a regression of $x_j$ on all other regressors $x_{j'\neq j}$.

## Implications of the $b_j$ Formula

Given our formula, we can see that:

1) $\tilde{x}_{ji}=x_{ji}-\bar{x}_{ji}$ if $x_j$ is independent of all other regressors

2) $\tilde{x}_{ji}=x_{ji}-\bar{x}_{ji}$ if $x_j$ is our only regressor (*bivariate regression*)

3) $\tilde{x}_{ji}=0$ if $x_j$ is perfectly colinear (exactly overlaps) with one or more other regressor $x_{j'\neq j}$ in our model

In cases (1) and (2), we see that $b_j=b_j^{\text{bivariate}}$ when we plug in $\tilde{x}_j$.

In case (3), we see that $\tilde{x}_j=0$ will mean *we cannot estimate* $b_2$ as we will have a zero in our denominator.

## Overlapping Variation: Diagram

Variation of $x_j$ overlapping with other regressors:

```{r}
#| label: fig-venn-overlap
#| fig-height: 2.6
#| fig-width: 5.5
#| fig-pos: "H"
#| fig-cap: "Total variation of each regressor (full circle) and the unique variation used for OLS estimation (shaded non-overlapping region)."
#| fig-alt: "Schematic Venn diagram with three overlapping circles illustrating regressor variation in multivariate OLS. The top circle has a blue border and is labeled x_2 above it. The bottom-left circle has a purple border and is labeled x_3 to its left. The bottom-right circle has a red border and is labeled x_4 to its right. Each circle represents the total variation of one regressor. Only the non-overlapping portion of each circle is shaded, labeled Unique variation outside the circle. The unshaded overlapping regions represent shared variation that cannot be uniquely attributed to any one regressor; an arrow from outside the diagram points to the central overlap region, labeled Shared variation."
library(ggplot2)

r   <- 1.0
cx  <- c( 0.00, -0.85,  0.85)
cy  <- c( 0.75, -0.45, -0.45)

# Fine grid: shade only the non-overlapping (unique) portion of each circle
n   <- 400
xg  <- seq(-2.6, 2.6,   length.out = n)
yg  <- seq(-1.55, 2.35, length.out = n)
grd <- expand.grid(x = xg, y = yg)

in2 <- (grd$x - cx[1])^2 + (grd$y - cy[1])^2 <= r^2
in3 <- (grd$x - cx[2])^2 + (grd$y - cy[2])^2 <= r^2
in4 <- (grd$x - cx[3])^2 + (grd$y - cy[3])^2 <= r^2

grd$fill <- NA_character_
grd$fill[in2 & !in3 & !in4] <- "x2"
grd$fill[in3 & !in2 & !in4] <- "x3"
grd$fill[in4 & !in2 & !in3] <- "x4"

# Circle outlines
make_circle <- function(x0, y0, r, grp, n = 300) {
  theta <- seq(0, 2 * pi, length.out = n + 1)
  data.frame(x = x0 + r * cos(theta), y = y0 + r * sin(theta), grp = grp)
}
circ <- rbind(
  make_circle(cx[1], cy[1], r, "x2"),
  make_circle(cx[2], cy[2], r, "x3"),
  make_circle(cx[3], cy[3], r, "x4")
)

ggplot() +
  geom_raster(data = grd, aes(x = x, y = y, fill = fill), alpha = 0.30) +
  geom_path(data = circ, aes(x = x, y = y, group = grp, color = grp),
            linewidth = 1.2) +
  scale_fill_manual(values   = c(x2 = "#003087", x3 = "#ac47be", x4 = "#c0392b"),
                    na.value = "transparent") +
  scale_color_manual(values  = c(x2 = "#003087", x3 = "#ac47be", x4 = "#c0392b")) +
  # Variable name labels — outside each circle
  annotate("text", x =  0.00, y =  2.12, size = 3.5, fontface = "bold",
           color = "#003087", label = expression(italic(x)[2])) +
  annotate("text", x = -2.10, y = -0.15, size = 3.5, fontface = "bold",
           color = "#ac47be", label = expression(italic(x)[3])) +
  annotate("text", x =  2.10, y = -0.15, size = 3.5, fontface = "bold",
           color = "#c0392b", label = expression(italic(x)[4])) +
  # "Unique variation" labels — inside each circle's non-overlapping region
  annotate("text", x =  0.00, y =  1.38, size = 2.8, color = "#333333",
           label = "Unique\nvariation", lineheight = 0.85) +
  annotate("text", x = -1.05, y = -0.78, size = 2.8, color = "#333333",
           label = "Unique\nvariation", lineheight = 0.85) +
  annotate("text", x =  1.05, y = -0.78, size = 2.8, color = "#333333",
           label = "Unique\nvariation", lineheight = 0.85) +
  # "Shared variation" — outside all circles, arrow pointing to triple-overlap center
  annotate("text", x =  0.00, y = -1.55, size = 2.8, color = "#333333",
           label = "Shared\nvariation", lineheight = 0.85) +
  annotate("segment", x = 0, xend = 0, y = -1.38, yend = -0.45,
           arrow = arrow(length = unit(0.07, "inches"), type = "closed"),
           color = "#333333", linewidth = 0.5) +
  coord_equal(xlim = c(-2.6, 2.6), ylim = c(-1.75, 2.35)) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text       = element_blank(),
    axis.ticks      = element_blank(),
    axis.line       = element_blank()
  ) +
  labs(x = "", y = "")
```

## Why Unique Variation Matters

Why do we need to identify unique variation of $x_j$, $\tilde{x}_j$?

Suppose we want to estimate the associated change in $y$ when $x_2$ changes: this is our coefficient $b_2$. However, we also know that $x_2$ and $x_3$ are correlated and both appear in our regression^[$y=b_1+b_2x_2+b_3x_3+e$].

Now, suppose we observe that:

- $x_2$ increases

- $x_3$ increases

- $y$ increases

Did $y$ increase because $x_2$ changed? Or did $y$ increase because $x_3$ changed and $x_2$ changed *because it is correlated with* $x_3$? If variation in $x_2$ and $x_3$ (or any regressors) is allowed to overlap, we cannot identify the *unique* contribution $x_2$ makes to changes in $y$.

## Example: Estimating Coefficients in Practice

What would this look like in practice? Suppose we want to estimate $$\widehat{final}_i=b_1+b_2MT+b_3StataQuiz+b_4HWavg+b_5KCavg$$ for your grade on the final as a function of your other grades in this class. How would we estimate $b_3$, the coefficient on *StataQuiz*?

First, we would regress *StataQuiz* scores on all other regressors. Then, we would find the residuals of this regression. Finally, we would compute the ratio of the covariance of $x_3$ residuals with the residuals from our main regression to the variance of $x_3$ residuals.

## Partial Effect vs. Total Effect: Definitions

There are two effects from our regressor $b_j$ we might be interested in: the **partial effect** or the **total effect**

- The *partial effect* is the impact on our outcome variable when $x_j$ changes and *all other regressors are held constant*.

- The *total effect* is the impact on our outcome variable when $x_j$ changes and we allow all other regressors $x_{j'\neq j}$ that may covary with $x_j$ to change and impact our outcome variable through their associated coefficients.

It follows that the partial effect will be equal to the total effect if our regressor $x_j$ is *independent from all other regressors* $x_{j'\neq j}$.

## Partial and Total Effects: Formulas

We express the **partial effect** as:
$$\frac{\partial\hat{y}}{\partial x_j}=b_j$$

We express the **total effect** as:
$$\begin{align*}
\frac{d\hat{y}}{d x_j} &= \underbrace{\frac{\partial\hat{y}}{\partial x_j}}_{b_j}\underbrace{\frac{\partial x_j}{\partial x_j}}_1+\sum_{j'\neq j}\underbrace{\frac{\partial\hat{y}}{\partial x_{j'}}}_{b_{j'}}\frac{\partial x_{j'}}{\partial x_j} \\
&= \underbrace{b_j}_{\text{direct effect}}+\underbrace{\sum_{j'\neq j}b_{j'}\frac{\partial x_{j'}}{\partial x_j}}_{\text{indirect effect(s)}}
\end{align*}$$

## Total Effect Equals Bivariate Coefficient

With some math, we can also show that the *total effect* of $x_j$ in a multivariate regression of y is equal to the $b_2$ coefficient from a bivariate regression of y on $x_j$.

This should make sense: the total effect is the direct effect of $x_j$ on y plus any indirect effects through other variables. The sum of these effects must then be equal to the effect of $x_j$ on y alone in a bivariate regression.

## Model Fit in Multivariate Regression

Luckily, our measures of model fit do not need to change much to accommodate multivariate regression, since they depend on our residual ($e=y-\hat{y}$):

- RMSE $$s_e=\sqrt{\frac{1}{\boldsymbol{n-k}}\sum_{i=1}^n(y_i-\hat{y}_i)^2}$$

- $R^2$ $$\frac{ExpSS}{TSS}=\frac{\sum_{i=1}^n(\hat{y}_i-\bar{y})^2}{\sum_{i=1}^n(y_i-\bar{y})^2}$$

Notice: we now have $n-k$ degrees of freedom in our RMSE denominator, since we are estimating coefficients on $k$ regressors (instead of two) to use in our calculation.

## The Problem with $R^2$

Unhelpfully, our $R^2$ will always weakly increase as we add regressors^[Since $\sum e_i^2$ will never increase as more regressors are added, R-squared will either remain unchanged or increase as more regressors are added.], but we may face other downsides to endlessly adding regressors to our model.

We therefore might want to introduce a measure of model fit that:

1) Allows us to compare standardized fit between models

2) Penalizes models that endlessly add (useless) regressors

## Adjusted $R^2$

Thus, we introduce **adjusted R-squared**:

$$\bar{R}^2=1-\frac{ResSS/(n-k)}{TSS/(n-1)}\in[0,1]$$

This statistic still gives us a standardized measure of model fit but also *is decreasing in the number of regressors we add* ($k$). We typically prefer to use this statistic for multivariate regressions over the standard $R^2$.

## Multivariate Regression Example {.smaller shrink=30}

\vspace{3em}

```{stata}
#| label: reg-auto
qui sysuse auto
reg price mpg weight rep78
```

$$\widehat{price}=\underbrace{-1939.87}_{b_1}-\underbrace{52.22}_{b_2}\underbrace{mpg}_{x_2}+\underbrace{2.11}_{b_3}\underbrace{weight}_{x_3}+\underbrace{820.81}_{b_4}\underbrace{rep78}_{x_4}$$

# End of Lecture Material

## Knowledge Check 10

1) What is the difference between R-squared and adjusted R-squared? Why might we prefer the second as a measure of multivariate model fit?

2) Why do we have $n-k$ degrees of freedom in our multivariate RMSE instead of $n-2$?

3) How do we compute the coefficient $b_j$ in a multivariate regression with $k>2$ regressors (in words)?

4) How does the partial effect of $x_j$ differ from the total effect?
