---
title: "ECN 102: Analysis of Economics Data"
subtitle: "Chapter 12: Errors and Prediction"
author: "Remy Beauregard"
lang: en
format:
  revealjs:
    output-file: "ECN 102 Lecture 12.html"
    slide-level: 2
    theme: simple
    transition: none
    smaller: true
    include-in-header:
      text: |
        <meta name="description" content="Lecture slides 12 for ECN 102.">
  beamer:
    output-file: "ECN 102 Lecture 12.pdf"
  html:
    output-file: "ECN 102 Lecture 12 Plain.html"
    include-in-header:
      text: |
        <meta name="description" content="Lecture slides 12 for ECN 102.">
execute:
  echo: false
---

```{r}
#| label: setup
#| include: false
require("Statamarkdown")
```

## Standard Errors: Overview

Our standard error for a statistic is our best estimate for the population standard deviation of that statistic. From this measure, we derive our inference about a population parameter from our sample estimation. Thus, it is critical that we are using the proper standard error for a given setting. We will think about a few cases where typical standard errors would be inappropriate and would lead to improper statistical inference.

## Robust Standard Errors

We have discussed the first case already: **robust standard errors**. Just like in the bivariate case, we invoke robust standard errors (with the `robust` option in Stata) to address possible threats to homoskedasticity. However, it is common practice in Economics to default to using robust standard errors, since our assumption of homoskedasticity frequently is violated.

We do not worry about over-using robust standard errors, as these errors are valid even if homoskedasticity *does* hold. However, if it does not hold, non-robust standard errors would be incorrect. Our use of robust standard errors relaxes our third assumption to be:

$$Var[u_i|x_{2i},...,x_{ki}]=\sigma^2_{ui},\hspace{5pt} i=1,...n$$

## Clustered Standard Errors

In addition to being heteroskedastic, we may think that our errors may be correlated for given sets of observations and violate independence. In this case, we would want to **cluster** our observations by some group variable $G$ to address this threat to the fourth assumption.

For example, if we are studying the effect of a farm subsidy program on household crop outcomes, we may anticipate that harvests would be correlated for all households in a given village. In this case, we would want to cluster our standard errors at the village level, to account for the fact that:

1. We assume errors *within* a cluster will be correlated<br>(not independent)

2. We assume errors *between* clusters will not be correlated (independent)

## Clustered Standard Errors: Modified Assumptions

Clustering our standard errors *automatically uses robust standard errors* and changes our four population assumptions to account for the fact that errors within clusters may be correlated but errors between clusters should not be:

$2_{clu}$: Cluster unbiasedness
$$E[u_i|x_{2i},...x_{ki},x_{2j},...,x_{kj}]=0\text{ for }i\text{ and }j\text{ in the same cluster}$$

$3_{clu}$: Cluster homoskedasticity
$$Var[u_i|x_{2i},...,x_{ki}]=\sigma^2_{ui}\text{ for }i=1,...,N$$

$4_{clu}$: Cluster independence
$$Cor[u_i,u_j|x_{2i},...,x_{ki},x_{2j},...,x_{kj}]
\begin{cases}
\neq 0 & i\text{ and }j\text{ in the same cluster}\\
= 0 & i\text{ and }j\text{ in different clusters}
\end{cases}$$

## Bootstrapped Standard Errors

Finally, we may suspect that our data does not even fulfill linearity ($\hat{y}=\beta_1+\beta_2x_2+...+\beta_kx_k$). In this case, our estimated standard errors for $b_1,b_2,...,b_k$ may be too large and we need to **bootstrap**.

Bootstrapping is essentially treating our sample as the population and resampling (with replacement) a large number of times to compute new statistics. Then, the variation across our different bootstrap samples gives us a "bootstrapped" estimate of our standard error for that statistic.

## Bootstrap: Example Procedure

For example, if we want to bootstrap our variance of $b_2$ for a sample $n=5$ of $(x,y)$, we would:

1. Draw $S$ samples of size $n=5$ from our "population" of data with replacement

2. Compute our $b_2^*$ for each of these samples and compute the mean across these samples, $\bar{b_2^*}$

3. Estimate our variance with $\hat{V}_{boot}[b_2]=\frac{1}{S-1}\sum_{s=1}^S\left(b_{2_s}^*-\bar{b_2^*}\right)^2$

If our population model was quadratic e.g. $y=\beta_1+\beta_2x+\beta_3x^2+u$ we would expect to find $\hat{V}_{boot}[b_2]<s^2_{b_2}$ indicating our regular OLS standard errors were overestimating the variability of $b_2$ due to our model being ill-fitting. We should therefore adjust our regression.

## Choosing the Right Standard Error

**Do you suspect...**

- *Heteroskedasticity*? $\Rightarrow$ robust standard errors

  `reg y x, robust`

- *Error dependence within clusters*? $\Rightarrow$ clustered standard errors

  `reg y x, vce(cluster GROUPVAR)`

- *Non-linearity*? $\Rightarrow$ bootstrapped standard errors

  `bootstrap, reps(S): reg y x`

## Prediction: Average vs. Individual Values

We may want to use our regression to predict both average values and individual values of our outcome based on our regressors. Average values will be the conditional mean of y, $E[y|x_2,...,x_k]$, while actual values will be the conditional value of y, $y|x_2,...,x_k$. It should be clear that the second quantity will be more widely dispersed than the first; it represents the full range of outcome values we might expect, not just the average values in the middle.

In both cases, we would use $y|x^*_2,...,x_k^*$ for a given $x=x^*$ by multiplying our regressor values by our estimated coefficients, but we would also need to consider the addition of the error term $u$ for our individual values. We do not know very much about this error (besides it being conditional mean zero under assumption 2) but it will affect how spread out our predicted individual values will be.

## Power: Definition

Previously, we defined our *test size* as the $\Pr(\text{Type I})=\alpha$ but left a discussion of our Type II error probability ambiguous. Now, we can more precisely define the **power** of our test as $1-\Pr(\text{Type II})$, which we seek to maximize for a given test size.

If we are able to choose our sample size, we may need to calculate an $n$ such that our results will be **well-powered** - our standard errors will not be enormous and our results will be believable. Of course, it is often very expensive to increase our sample size, so we balance constraints on sampling with concerns about power. As before, we further cement the power of our tests by using our *best estimators*, or those with minimum variance in their class, to reduce the chances of a Type II error as much as possible.

## Power: Type II Error Regions

:::: {.columns}
::: {.column width="50%"}
Large type II error region:

```{r}
#| label: fig-typeII-large
#| fig-height: 2.4
#| fig-width: 4.5
#| fig-cap: "Large type II error region when the alternative is close to the null."
#| fig-alt: "Schematic density plot illustrating a large type II error region. The null distribution H0 is a solid dark-blue bell curve centered at zero. The alternative distribution HA is a dashed purple bell curve centered slightly to the right, heavily overlapping H0. A vertical dotted line marks the rejection threshold. The large purple-shaded area under HA to the left of the threshold is the type II error region, labeled Type II. The small blue-shaded area under H0 to the right of the threshold is labeled alpha. When the alternative is close to the null, most of HA falls in the fail-to-reject zone, so power is low."
#| warning: false
library(ggplot2)
set.seed(42)
crit <- qnorm(0.95)
x    <- seq(-3.5, 5.0, length.out = 600)
d1   <- data.frame(x = x, h0 = dnorm(x, 0, 1), ha = dnorm(x, 1.0, 1))
ggplot(d1, aes(x = x)) +
  geom_area(data = subset(d1, x >= crit), aes(y = h0),
            fill = "#003087", alpha = 0.20) +
  geom_area(data = subset(d1, x <= crit), aes(y = ha),
            fill = "#ac47be", alpha = 0.25) +
  geom_line(aes(y = h0), color = "#003087", linetype = "solid",  linewidth = 0.7) +
  geom_line(aes(y = ha), color = "#ac47be", linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = crit, color = "#333333", linetype = "dotted", linewidth = 0.5) +
  annotate("text", x = -1.2, y = 0.30, label = expression(H[0]),
           color = "#003087", size = 3.0, hjust = 0.5) +
  annotate("text", x =  2.1, y = 0.35, label = expression(H[A]),
           color = "#ac47be", size = 3.0, hjust = 0.5) +
  annotate("text", x = 0.4, y = 0.06, label = "Type II",
           color = "#ac47be", size = 2.8, hjust = 0.5) +
  annotate("text", x = 2.0, y = 0.025, label = expression(alpha),
           color = "#333333", size = 2.8, hjust = 0.5) +
  labs(x = "Test statistic", y = "Density") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text  = element_blank(),
        axis.ticks = element_blank())
```
:::
::: {.column width="50%"}
Small type II error region:

```{r}
#| label: fig-typeII-small
#| fig-height: 2.4
#| fig-width: 4.5
#| fig-cap: "Small type II error region when the alternative is far from the null."
#| fig-alt: "Schematic density plot illustrating a small type II error region. The null distribution H0 is a solid dark-blue bell curve centered at zero. The alternative distribution HA is a dashed purple bell curve centered far to the right, with almost no overlap with H0. A vertical dotted line marks the rejection threshold. The small purple-shaded sliver under HA to the left of the threshold is labeled Type II with an arrow. Almost all of the HA distribution lies to the right of the threshold, indicating high power."
#| warning: false
library(ggplot2)
d2 <- data.frame(x = x, h0 = dnorm(x, 0, 1), ha = dnorm(x, 2.8, 1))
ggplot(d2, aes(x = x)) +
  geom_area(data = subset(d2, x >= crit), aes(y = h0),
            fill = "#003087", alpha = 0.20) +
  geom_area(data = subset(d2, x <= crit), aes(y = ha),
            fill = "#ac47be", alpha = 0.25) +
  geom_line(aes(y = h0), color = "#003087", linetype = "solid",  linewidth = 0.7) +
  geom_line(aes(y = ha), color = "#ac47be", linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = crit, color = "#333333", linetype = "dotted", linewidth = 0.5) +
  annotate("text", x = -1.2, y = 0.30, label = expression(H[0]),
           color = "#003087", size = 3.0, hjust = 0.5) +
  annotate("text", x =  2.1, y = 0.35, label = expression(H[A]),
           color = "#ac47be", size = 3.0, hjust = 0.5) +
  annotate("segment", x = 0.65, xend = 1.15, y = 0.07, yend = 0.02,
           color = "#ac47be", linewidth = 0.4,
           arrow = arrow(length = unit(0.08, "cm"), type = "closed")) +
  annotate("text", x = 0.45, y = 0.09, label = "Type II",
           color = "#ac47be", size = 2.8, hjust = 0.5) +
  annotate("text", x = 2.0, y = 0.025, label = expression(alpha),
           color = "#333333", size = 2.8, hjust = 0.5) +
  labs(x = "Test statistic", y = "Density") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text  = element_blank(),
        axis.ticks = element_blank())
```
:::
::::

## Power: Estimator Distribution

:::: {.columns}
::: {.column width="50%"}
Parameter assumption and "best estimator" distribution:

```{r}
#| label: fig-power-blue
#| fig-height: 2.4
#| fig-width: 4.5
#| fig-cap: "High power using the best linear unbiased estimator (minimum variance)."
#| fig-alt: "Schematic density plot illustrating high power from the best linear unbiased estimator. The null distribution H0 is a solid dark-blue bell curve centered at zero. The BLUE sampling distribution is a narrow dashed purple bell curve centered to the right of the rejection threshold at 1.645, reflecting its low variance and tall narrow shape. The small purple-shaded region under BLUE to the left of the threshold is labeled Type II. The large blue-shaded region to the right is labeled Power. Because BLUE has minimum variance, its distribution is concentrated to the right of the threshold, yielding high power."
#| warning: false
library(ggplot2)
xb <- seq(-3.5, 7.0, length.out = 600)
db <- data.frame(x = xb, h0 = dnorm(xb, 0, 1), ha = dnorm(xb, 2.4, 0.7))
ggplot(db, aes(x = x)) +
  geom_area(data = subset(db, x <= crit), aes(y = ha),
            fill = "#ac47be", alpha = 0.20) +
  geom_area(data = subset(db, x >= crit), aes(y = ha),
            fill = "#003087", alpha = 0.20) +
  geom_line(aes(y = h0), color = "#003087", linetype = "solid",  linewidth = 0.7) +
  geom_line(aes(y = ha), color = "#ac47be", linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = crit, color = "#333333", linetype = "dotted", linewidth = 0.5) +
  annotate("text", x = -1.2, y = 0.30, label = expression(H[0]),
           color = "#003087", size = 3.0, hjust = 0.5) +
  annotate("text", x =  2.4, y = 0.60, label = "BLUE",
           color = "#ac47be", size = 3.0, hjust = 0.5) +
  annotate("segment", x = 0.50, xend = 1.22, y = 0.11, yend = 0.04,
           color = "#ac47be", linewidth = 0.4,
           arrow = arrow(length = unit(0.08, "cm"), type = "closed")) +
  annotate("text", x = 0.30, y = 0.15, label = "Type II",
           color = "#ac47be", size = 2.8, hjust = 0.5) +
  annotate("text", x = 2.8, y = 0.09, label = "Power",
           color = "#003087", size = 2.8, hjust = 0.5) +
  labs(x = "Estimator value", y = "Density") +
  coord_cartesian(ylim = c(0, 0.65)) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text  = element_blank(),
        axis.ticks = element_blank())
```
:::
::: {.column width="50%"}
Parameter assumption and other estimator distribution:

```{r}
#| label: fig-power-other
#| fig-height: 2.4
#| fig-width: 4.5
#| fig-cap: "Lower power using a higher-variance linear unbiased estimator with the same center as BLUE."
#| fig-alt: "Schematic density plot illustrating low power from a higher-variance linear unbiased estimator. The null distribution H0 is a solid dark-blue bell curve centered at zero. The other linear unbiased estimator's sampling distribution is a wide dashed purple bell curve centered at the same true parameter value as BLUE, but with higher variance giving a shorter and wider shape. The large purple-shaded region to the left of the threshold is labeled Type II. The smaller blue-shaded region to the right is labeled Power. Because this estimator has higher variance than BLUE, its wider spread places more probability mass in the fail-to-reject region, yielding lower power despite having the same center."
#| warning: false
library(ggplot2)
xp <- seq(-3.5, 7.0, length.out = 600)
do <- data.frame(x = xp, h0 = dnorm(xp, 0, 1), ha = dnorm(xp, 2.4, 1.4))
ggplot(do, aes(x = x)) +
  geom_area(data = subset(do, x <= crit), aes(y = ha),
            fill = "#ac47be", alpha = 0.25) +
  geom_area(data = subset(do, x >= crit), aes(y = ha),
            fill = "#003087", alpha = 0.15) +
  geom_line(aes(y = h0), color = "#003087", linetype = "solid",  linewidth = 0.7) +
  geom_line(aes(y = ha), color = "#ac47be", linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = crit, color = "#333333", linetype = "dotted", linewidth = 0.5) +
  annotate("text", x = -1.2, y = 0.30, label = expression(H[0]),
           color = "#003087", size = 3.0, hjust = 0.5) +
  annotate("text", x = 1.8, y = 0.40, label = "Other linear unbiased estimator",
           color = "#ac47be", size = 2.6, hjust = 0.0) +
  annotate("text", x = 0.7, y = 0.05, label = "Type II",
           color = "#ac47be", size = 2.8, hjust = 0.5) +
  annotate("text", x = 3.5, y = 0.05, label = "Power",
           color = "#003087", size = 2.8, hjust = 0.5) +
  labs(x = "Estimator value", y = "Density") +
  coord_cartesian(ylim = c(0, 0.65)) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text  = element_blank(),
        axis.ticks = element_blank())
```
:::
::::

## Multiple Hypothesis Testing

**NOT ON EXAM**

For a single test, we know that $\Pr(\text{Type I})=\alpha$. If we ran 20 tests in a row, however, we would expect that (on average) one of those tests yielded a *spurious result* - a Type I error. Although we will not dig into the methodology in this class, we should consider additional measures to employ when we are doing many individual tests on the same sample of data to account for the fact that some of our results may be false positives.

These **multiple hypothesis corrections** will be used to show the robustness of our results when we have to run many tests and are worried that our probability of a type I error has compounded to a worryingly large amount.

# End of lecture material

## Knowledge Check 12

Consider a test for some parameter $\theta$ (e.g. $\beta_2$) with $$H_0:\theta=0$$ $$H_A:\theta\neq0$$

Suppose we know that the probability of rejecting $H_0$ is 0.10 when $\theta=0$ and 0.75 when $\theta\neq0$. Please give:

a) The size of the test

b) The probability of a type I error

c) The probability of a type II error

d) The power of the test

Suppose $n$ is fixed. What could we do to ensure our test is as well-powered as possible?
