---
title: "ECN 102: Analysis of Economics Data"
subtitle: "Chapter 14: Indicator Variables"
author: "Remy Beauregard"
lang: en
format:
  revealjs:
    output-file: "ECN 102 Lecture 14.html"
    slide-level: 2
    theme: simple
    transition: none
    smaller: true
    include-in-header:
      text: |
        <meta name="description" content="Lecture slides 14 for ECN 102.">
  beamer:
    output-file: "ECN 102 Lecture 14.pdf"
  html:
    output-file: "ECN 102 Lecture 14 Plain.html"
    include-in-header:
      text: |
        <meta name="description" content="Lecture slides 14 for ECN 102.">
execute:
  echo: false
---

```{r}
#| label: setup
#| include: false
require("Statamarkdown")
```

## Indicator Variables: Introduction

We previously asserted that categorical variables, such as "shirt color", are not typically included in regression, since it is difficult to interpret a numerical coefficient for a discrete categorical series.

One common exception is using **indicator variables** which encode categorical information. A special kind of indicator variable is called a **dummy variable** which are coded as being only zero or one.

We may have a single indicator variable for a binary outcome (*male*=1 vs. *male*=0) or a set of indicator variables that span the possible options (*red shirt, blue shirt, green shirt, ...*). Indicator variables give us a powerful way to analyze differences in means and slopes for different groups within our regression.

## Simple Dummy Variable Regression

Suppose we want to estimate the average final grade in ECN102 for Econ majors and non-majors. We could define a dummy variable $ECN=1$ for Economics majors and 0 otherwise.

We then regress $\widehat{final}=b_1+a_1\times ECN$:

$$\widehat{final}_i=\begin{cases}
b_1+a_1 & ECN_i =1 \\
b_1 & ECN_i=0
\end{cases}$$

Our coefficient $a_1$ on $ECN$ is then the average difference in final grade between Econ majors and non-majors. This is the simplest use of a dummy variable.

## Graph: Final Grade on ECN Major Status

$\widehat{final}=b_1+a_1 ECN$:

```{r}
#| label: fig-dummy-intercept-only
#| fig-height: 2.6
#| fig-width: 5.5
#| fig-cap: "Schematic regression lines showing the effect of a binary dummy variable on the outcome, with no continuous regressor."
#| fig-alt: "Schematic graph with Hours studying on the x-axis and Final grade on the y-axis, both axes unlabeled numerically. Two horizontal lines span the full x-range: a solid dark-blue line for Non-major (ECN = 0) at a lower level, and a dashed purple line for Econ major (ECN = 1) at a higher level. A double-headed vertical arrow between the two lines near x = 2.5 is labeled a1, indicating the intercept shift from the dummy variable. The graph is conceptual and shows that with only a dummy regressor the model predicts a constant average grade for each group, with the two groups differing only by a1."
library(ggplot2)
b1 <- 55; a1 <- 10

df1 <- data.frame(
  x     = c(0, 10, 0, 10),
  y     = c(b1, b1, b1 + a1, b1 + a1),
  group = c("nm", "nm", "em", "em")
)

ggplot(df1, aes(x, y, color = group, linetype = group)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("nm" = "#003087", "em" = "#ac47be")) +
  scale_linetype_manual(values = c("nm" = "solid", "em" = "dashed")) +
  annotate("segment", x = 2.5, xend = 2.5, y = b1, yend = b1 + a1,
           arrow = arrow(ends = "both", length = unit(0.08, "inches")),
           color = "#333333", linewidth = 0.5) +
  annotate("text", x = 3.1, y = b1 + a1 / 2,
           label = expression(italic(a)[1]),
           color = "#333333", size = 3.2, hjust = 0) +
  annotate("text", x = 7, y = b1 - 3.5,
           label = "Non-major (ECN = 0)", color = "#003087", size = 3.0, hjust = 0.5) +
  annotate("text", x = 7, y = b1 + a1 + 3.5,
           label = "Econ major (ECN = 1)", color = "#ac47be", size = 3.0, hjust = 0.5) +
  coord_cartesian(xlim = c(0, 10), ylim = c(45, 100)) +
  labs(x = "Hours studying", y = "Final grade") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Dummy Variables in Multivariate Regression

We may also want to include dummy variables in a multivariate regression with other continuous numerical regressors. In this case, the partial effect (estimated coefficient) of the dummy variable is the change in our outcome when our dummy goes from 0 to 1, all other regressors held constant. The total effect is our partial effect added to the product of any correlated movement in other variables and its associated coefficient.

We now run the regression: $$\widehat{final}_i=b_1+b_2hrs+a_1 ECN$$ where $hrs$ is the number of hours spent studying and $ECN=1$ is being an Econ major. Now, our estimated coefficient $a_1$ is the average difference in final grades between Econ majors and non-majors holding time spent studying fixed.

## Graph: Final Grade, Hours, and ECN Major Status

$\widehat{final}=b_1+b_2hrs+a_1 ECN$:

```{r}
#| label: fig-dummy-parallel
#| fig-height: 2.6
#| fig-width: 5.5
#| fig-cap: "Schematic parallel regression lines showing a binary dummy variable added to a regression with a continuous regressor."
#| fig-alt: "Schematic graph with Hours studying on the x-axis and Final grade on the y-axis, both unlabeled numerically and on the same scale as the preceding graph. Two upward-sloping lines with equal slope b2 run in parallel: a solid dark-blue line for Non-major (ECN = 0) at a lower level and a dashed purple line for Econ major (ECN = 1) at a higher level. A double-headed vertical arrow near x = 1.5 is labeled a1, showing that the constant vertical gap between the two lines equals a1 regardless of hours studied. The parallel slopes illustrate that b2 is the same for both groups when there is no interaction term."
library(ggplot2)
b1 <- 55; b2 <- 2; a1 <- 10

df2 <- data.frame(
  x     = c(0, 10, 0, 10),
  y     = c(b1, b1 + b2 * 10, b1 + a1, b1 + a1 + b2 * 10),
  group = c("nm", "nm", "em", "em")
)

ggplot(df2, aes(x, y, color = group, linetype = group)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("nm" = "#003087", "em" = "#ac47be")) +
  scale_linetype_manual(values = c("nm" = "solid", "em" = "dashed")) +
  annotate("segment", x = 1.5, xend = 1.5,
           y = b1 + b2 * 1.5, yend = b1 + a1 + b2 * 1.5,
           arrow = arrow(ends = "both", length = unit(0.08, "inches")),
           color = "#333333", linewidth = 0.5) +
  annotate("text", x = 2.1, y = b1 + b2 * 1.5 + a1 / 2,
           label = expression(italic(a)[1]),
           color = "#333333", size = 3.2, hjust = 0) +
  annotate("text", x = 7, y = b1 + b2 * 7 - 3.5,
           label = "Non-major (ECN = 0)", color = "#003087", size = 3.0, hjust = 0.5) +
  annotate("text", x = 7, y = b1 + a1 + b2 * 7 + 3.5,
           label = "Econ major (ECN = 1)", color = "#ac47be", size = 3.0, hjust = 0.5) +
  coord_cartesian(xlim = c(0, 10), ylim = c(45, 100)) +
  labs(x = "Hours studying", y = "Final grade") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Interacting a Dummy Variable with a Regressor

Finally, we could **interact** our dummy variable with another regressor to capture the average difference in *slopes* between one category and another. Adding to our regression of final grades, we could regress $$\widehat{final}_i=b_1+b_2hrs+a_1ECN+a_2(ECN\times hrs)$$

Now:

- $b_2$ captures the change in final grade associated with 1 additional hour of studying for *non*-majors

- $a_1$ captures the average difference in overall final grades between majors and non-majors

- $b_2+a_2$ captures the change in final grade associated with 1 additional hour of studying for *majors*

## Graph: Interacted Dummy Regression

$\widehat{final}=b_1+b_2hrs+a_1ECN+a_2(ECN\times hrs)$:

```{r}
#| label: fig-dummy-interaction
#| fig-height: 2.6
#| fig-width: 5.5
#| fig-cap: "Schematic regression lines showing a binary dummy variable interacted with a continuous regressor."
#| fig-alt: "Schematic graph with Hours studying on the x-axis and Final grade on the y-axis, both unlabeled numerically and on the same scale as the preceding graphs. Two upward-sloping lines diverge as hours increase: a solid dark-blue line for Non-major (ECN = 0) with a shallower slope b2, and a dashed purple line for Econ major (ECN = 1) with a steeper slope b2 plus a2. A double-headed vertical arrow near the left axis labeled a1 marks the initial intercept gap between the two groups at zero hours. As hours increase the gap between the lines grows because a2 is positive, illustrating how the interaction term allows the return to studying to differ between the two groups."
library(ggplot2)
b1 <- 55; b2 <- 2; a1 <- 10; a2 <- 1.2

df3 <- data.frame(
  x     = c(0, 10, 0, 10),
  y     = c(b1, b1 + b2 * 10, b1 + a1, b1 + a1 + (b2 + a2) * 10),
  group = c("nm", "nm", "em", "em")
)

ggplot(df3, aes(x, y, color = group, linetype = group)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("nm" = "#003087", "em" = "#ac47be")) +
  scale_linetype_manual(values = c("nm" = "solid", "em" = "dashed")) +
  annotate("segment", x = 0.5, xend = 0.5, y = b1, yend = b1 + a1,
           arrow = arrow(ends = "both", length = unit(0.08, "inches")),
           color = "#333333", linewidth = 0.5) +
  annotate("text", x = 1.1, y = b1 + a1 / 2,
           label = expression(italic(a)[1]),
           color = "#333333", size = 3.2, hjust = 0) +
  annotate("text", x = 7, y = b1 + b2 * 7 - 3.5,
           label = "Non-major (ECN = 0)", color = "#003087", size = 3.0, hjust = 0.5) +
  annotate("text", x = 7, y = b1 + a1 + (b2 + a2) * 7 + 3.5,
           label = "Econ major (ECN = 1)", color = "#ac47be", size = 3.0, hjust = 0.5) +
  coord_cartesian(xlim = c(0, 10), ylim = c(45, 100)) +
  labs(x = "Hours studying", y = "Final grade") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Interpreting the Interacted Dummy: Slopes and Intercepts

For an interacted dummy variable regression: $$\hat{y}=b_1+b_2x+a_1d+a_2(d\times x)$$

- $a_1>0$ means the intercept for $d=1$ is *higher* than for $d=0$

- $a_1<0$ means the intercept for $d=1$ is *lower* than for $d=0$

- $a_2>0$ means the slope for $d_1$ is *steeper* than for $d=0$

- $a_2<0$ means the slope for $d_1$ is *less steep* than for $d=0$

If $a_2$ is larger than $b_2$ in magnitude but has the opposite sign, the slope for $d=1$ will go in the *opposite direction* than for $d=0$.

## Interacted Dummy: Two Separate Group Regressions

Plugging in $d=0$ and $d=1$, we see that we are actually just running two different bivariate regressions for two groups:

$$\begin{align*}
d=0\text{: }\hat{y}&=b_1+b_2x+a_1d+a_2(d\cdot x)\\
&=b_1+b_2x+a_1(0)+a_2(0)x\\
&=b_1+b_2x
\end{align*}$$

$$\begin{align*}
d=1\text{: }\hat{y}&=b_1+b_2x+a_1d+a_2(d\cdot x)\\
&=b_1+b_2x+a_1(1)+a_2(1)x\\
&=b_1+b_2x+a_1+a_2x\\
&= \underbrace{(a_1+b_1)}_{c_1} + \underbrace{(a_2+b_2)}_{c_2}x\\
&=c_1+c_2x
\end{align*}$$

## Example: Car Prices, MPG, and Country of Origin {.smaller shrink=20}

How would we interpret the following regression results with indicator variable *foreign*?

```{stata}
#| label: load-auto-l14
#| collectcode: true
#| results: false
qui sysuse auto
g mpgXforeign = mpg*foreign
```

```{stata}
#| label: reg-price-mpg-foreign
reg price mpg foreign mpgXforeign
```

$$price=\beta_1+\beta_2mpg+\alpha_1foreign+\alpha_2(mpg\times foreign)+u$$

## Graph: Price on MPG by Country of Origin

Graph of $price$ on $mpg$ across $foreign$:

## Ordered and Unordered Categorical Variables

We may use indicator variables that take values other than 0 or 1 if we have an *ordered* or *ranked* categorical variable, such as class year (freshman, sophomore, etc.). We could capture these with a single variable called $Class$ which takes values 0, 1, 2, or 3 depending on the incoming year (with 4 for super-senior+). Our only caveat in this case would be that we are assuming the effect of going from freshman to sophomore is the same as going from junior to senior.

For *unordered* categorical variables, we may instead want to use a set of dummy variables, each representing a category in our set, such as shirt colors. We say these dummies are **mutually exclusive** and **exhaustive** if all observations fall into exactly one category. We can check this by adding the sum of the sample means for our dummies — they should sum to 1 exactly.

## The Dummy Variable Trap

We previously discussed the issue of *perfect collinearity* between regressors — we cannot estimate two separate regressors that are exact linear combinations of each other. For dummy variables, this means we cannot include dummy variables for all categories in our regression without errors — we must leave out one variable. We call this group, unsurprisingly, the **leave-out group**, and all regression results are compared to this group. We may want to choose this group carefully to yield the best interpretation of our results.

Failing to remove one mutually exclusive dummy variable from a set that spans all categories is called the **dummy variable trap** and will lead to errors in estimation. We can also avoid this by including all dummies but *dropping the constant term*.

## Why the Dummy Variable Trap Occurs

Why do we encounter the dummy variable trap?

Suppose we have a category $c$ that takes values 1, 2, or 3 and every observation falls into exactly one category. Now suppose we construct dummy variables $d_1,d_2,d_3$ for if an observation falls into category 1, 2, or 3, respectively.

If $d_1=0$ and $d_2=0$, it **must** be the case that $d_3=1$.^[Or if $d_1=1$ or $d_2=1$, it must be the case that $d_3=0$.]

We can then rewrite $d_3$ as $d_3=1-d_1-d_2$. This means $d_3$ is a *linear combination* of other regressors in our model and thus *perfectly collinear*. If we try to include $d_3$ in our model along with $d_1$ and $d_2$, we will not be able to estimate our coefficients.

## Rule: At Most n−1 Dummy Variables

Generally, if we have $n$ mutually exclusive and exhaustive categories each with a dummy variable, we can include at most $n-1$ dummy variables in our regression ($d_1,...,d_{n-1}$).

The last dummy variable, $d_n$, would be equal to: $$d_n=1-\sum_{i=1}^{n-1}d_i$$

More generally, if we choose the dummy variable for category $j$ to be our leave-out group, we will include all dummies $d_{i\neq j}$ in our regression and have $d_j=1-\sum_{i\neq j}d_i$.

## Inference on Dummy Variables

When we conduct inference on dummy variables with a leave-out group, we are not testing whether the included dummy is different from zero, but rather *if the difference between that group and the leave-out group* is zero. Our F-test for overall significance, however, still measures whether all *included* regressors meaningfully predict our outcome variable.

## Example: Values of the Foreign Variable {.smaller shrink=20}

In our car example, our variable *foreign* takes the following values:

```{stata}
#| label: fre-foreign
fre foreign
```

To estimate a regression using dummies for the values of *foreign*, we would need to either:

1. Include only a dummy on foreign and not domestic

2. Include only a dummy on domestic and not foreign

3. Drop our constant term and include both dummies

## Avoiding the Dummy Variable Trap {.smaller shrink=20}

Do we run into the dummy variable trap here? Why or why not?

```{stata}
#| label: reg-domestic-foreign
g domestic = foreign!=1
reg price mpg domestic foreign, nocons // remove constant term
```

$$\widehat{price}=\beta_2mpg+\delta_1domestic+\delta_2foreign$$

## Creating Factor Variable Dummies in Stata

We can easily create dummies for these discrete mutually exclusive **factor variables** in Stata:

```{stata}
#| label: ex-factor-gen
#| results: false
#| echo: true
#| eval: false
tab shirtcolor, g(dcolor) // generate dummies
reg price dcolor1 dcolor2 ... // dummy regression
```

We can also call on dummy variables directly in our regression. However, Stata makes us first convert categorical variables into numeric variables before we do this (*R* does not):

```{stata}
#| label: ex-factor-encode
#| results: false
#| echo: true
#| eval: false
encode shirtcolor, g(dcolor) // conversion
reg price i.dcolor // regression with factor variable
```

# End of Lecture Material

## Knowledge Check 14

Suppose we estimate the following regression: $$y = \beta_1 + \beta_2x + \alpha_1d + \alpha_2(d\times x)+ u$$ where $d$ is a dummy variable for being in category $C$ or not. How would we interpret the coefficients $b_1,b_2,a_1,$ and $a_2$?
